---
title: "Univariate TS Models (ARIMA/SARIMA)"
code-fold: true
execute:
  cache: true
  freeze: auto
---

## Model Setup

We load daily data for Bitcoin, S&P 500, NASDAQ, VIX, and the USD Index from FRED and merge the valid series into a single cleaned dataset. Returns are computed using first-differenced log prices.

::: {.panel-tabset}

### Data Preparation
```{r}
#| warning: false
#| message: false

library(quantmod)
library(tidyverse)
library(zoo)
library(forecast)
library(tseries)

start_date <- as.Date("2019-01-01")
end_date   <- as.Date("2025-09-18")

load_fred_data <- function(symbol) {
  tryCatch(
    getSymbols(symbol, src = "FRED", from = start_date, to = end_date,
               auto.assign = FALSE),
    error = function(e) { cat("FAILED:", symbol, "\n"); NULL }
  )
}

sp500_data  <- load_fred_data("SP500")
vix_data    <- load_fred_data("VIXCLS")
btc_data    <- load_fred_data("CBBTCUSD")
nasdaq_data <- load_fred_data("NASDAQCOM")
usd_data    <- load_fred_data("DTWEXBGS")

merge_list <- list(
  SP500   = sp500_data,
  VIX     = vix_data,
  Bitcoin = btc_data,
  NASDAQ  = nasdaq_data,
  USD     = usd_data
)
merge_list <- merge_list[!sapply(merge_list, is.null)]

price_zoo  <- do.call(merge.zoo, merge_list)
price_data <- fortify.zoo(price_zoo)
colnames(price_data) <- c("Date", names(merge_list))
price_data <- price_data |> na.omit()

get_series <- function(asset) price_data[[asset]]
get_diff_series <- function(asset) diff(log(price_data[[asset]])) |> na.omit()




cat(
  "\n=== DATASET SUMMARY ===\n",
  "Observations : ", nrow(price_data), "\n",
  "Assets Loaded: ", length(merge_list), "\n",
  "Date Range   : ", min(price_data$Date), " to ", max(price_data$Date), "\n"
)


```

```{r}
#| warning: false
#| message: false


assets <- names(merge_list)

arima_models <- list()
sarima_models <- list()

for (a in assets) {
series <- get_diff_series(a)

arima_models[[a]] <- auto.arima(series, stepwise=TRUE, approximation=TRUE)

sarima_models[[a]] <- auto.arima(series, seasonal=TRUE,
stepwise=TRUE, approximation=TRUE)
}


```



For modeling, each asset is fit with an ARIMA/SARIMA model using auto.arima() for efficient order selection. Models are stored once and reused throughout the page for diagnostics, ACF/PACF inspection, forecasting, and benchmark comparisons. 

### Helper Functions

```{r}
#| warning: false
#| message: false

plot_original_and_diff <- function(asset) {
  
  # Extract series
  original <- price_data[[asset]]
  dates    <- price_data$Date
  diff_ser <- get_diff_series(asset)
  
  # Layout
  par(mfrow = c(2, 1), mar = c(4, 4, 3, 1))
  
  # --- Original price plot ---
  plot(dates, original, type = "l", col = "steelblue",
       main = paste(asset, "- Original Series"),
       xlab = "Date", ylab = "Price Level")
  
  # --- First difference plot (log returns) ---
  plot(diff_ser, type = "l", col = "red",
       main = paste(asset, "- Log Returns (First Difference)"),
       xlab = "Index", ylab = "Log Return")
  
  # Reset layout
  par(mfrow = c(1, 1))
}


run_stationarity <- function(asset) {
print(adf.test(get_diff_series(asset)))
}

plot_acf_pacf <- function(asset) {
series <- get_diff_series(asset)
par(mfrow=c(1,2))
acf(series, main=paste(asset,"ACF"))
pacf(series, main=paste(asset,"PACF"))
par(mfrow=c(1,1))
}

run_arima_models <- function(asset) {
print(arima_models[[asset]])
}

run_arima_diagnostics <- function(asset) {
tsdiag(arima_models[[asset]])
}

forecast_arima_asset <- function(asset) {
plot(forecast(arima_models[[asset]], h=30),
main=paste(asset,"ARIMA Forecast"))
}

compare_benchmarks <- function(asset) {

  s <- get_diff_series(asset)

  f_naive  <- naive(s, h = 30)
  f_snaive <- snaive(s, h = 30)
  f_arima  <- forecast(arima_models[[asset]], h = 30)

  results <- tibble(
    Model = c("Naive", "Seasonal Naive", "ARIMA"),
    RMSE  = c(accuracy(f_naive)[,"RMSE"],
              accuracy(f_snaive)[,"RMSE"],
              accuracy(f_arima)[,"RMSE"]),
    MAE   = c(accuracy(f_naive)[,"MAE"],
              accuracy(f_snaive)[,"MAE"],
              accuracy(f_arima)[,"MAE"]),
    MAPE  = c(accuracy(f_naive)[,"MAPE"],
              accuracy(f_snaive)[,"MAPE"],
              accuracy(f_arima)[,"MAPE"])
  )

  print(results)

  # --------------------------------------------------------------
  # FIX: Extract safely and guarantee numeric, not NA, not ts
  # --------------------------------------------------------------
  arima_rmse <- suppressWarnings(as.numeric(results$RMSE[results$Model == "ARIMA"]))
  best_bench <- suppressWarnings(as.numeric(min(results$RMSE[results$Model != "ARIMA"], na.rm = TRUE)))

  # If still NA, replace with Inf so comparison will work
  if (is.na(arima_rmse)) arima_rmse <- Inf
  if (is.na(best_bench)) best_bench <- Inf

  pct_diff <- 100 * (best_bench - arima_rmse) / best_bench

  # --------------------------------------------------------------
  # Interpretation
  # --------------------------------------------------------------
  cat("\nInterpretation for", asset, "\n")

  if (isTRUE(arima_rmse < best_bench)) {

    cat(
      "ARIMA outperforms the benchmark models.\n",
      "It improves RMSE by", round(pct_diff, 2), "% relative to the best naive model.\n"
    )

  } else if (isTRUE(arima_rmse > best_bench)) {

    cat(
      "ARIMA does NOT beat the benchmarks.\n",
      "It performs", round(-pct_diff, 2), "% worse than the best naive model.\n"
    )

  } else {

    cat("ARIMA performs similarly to the naive benchmarks.\n")

  }

  invisible(results)
}



run_sarima_models <- function(asset) {
print(sarima_models[[asset]])
}

forecast_sarima_asset <- function(asset) {
plot(forecast(sarima_models[[asset]], h=30),
main=paste(asset,"SARIMA Forecast"))
}

seasonality_analysis <- function(asset) {
s <- ts(get_diff_series(asset), frequency=365)
plot(stl(s, "periodic"), main=paste(asset,"Seasonality"))
}

# FAST CV — no repeated auto.arima calls

cross_validate_sarima <- function(asset) {
series <- ts(get_diff_series(asset))
model <- sarima_models[[asset]]
ffun <- function(y, h) forecast(model, h=h)
e <- tsCV(series, ffun, h=1)
mean(e^2, na.rm=TRUE)
}

```

What These Helper Functions Do

The following helper functions streamline the ARIMA/SARIMA workflow by wrapping common analysis steps into reusable tools:

- **`plot_original_and_diff()`**  
  Plots the raw price series and its first-differenced log returns to visualize stationarity and volatility changes.

- **`run_stationarity()`**  
  Runs the Augmented Dickey–Fuller (ADF) test on the differenced series to confirm stationarity before modeling.

- **`plot_acf_pacf()`**  
  Produces ACF and PACF plots of the return series, helping identify potential ARIMA orders.

- **`run_arima_models()`**  
  Prints the fitted ARIMA model for a selected asset.

- **`run_arima_diagnostics()`**  
  Displays residual diagnostics for the ARIMA model (autocorrelation, Ljung–Box tests).

- **`forecast_arima_asset()`**  
  Generates a 30-day ARIMA forecast with confidence intervals.

- **`compare_benchmarks()`**  
  Compares ARIMA forecasts to simple alternatives: Naive and Seasonal Naive models.

- **`run_sarima_models()`**  
  Prints the asset’s fitted SARIMA model (seasonal version).

- **`forecast_sarima_asset()`**  
  Produces a 30-day SARIMA forecast for comparison with ARIMA.

- **`seasonality_analysis()`**  
  Performs STL decomposition to examine seasonal effects in the differenced series.

- **`cross_validate_sarima()`**  
  Performs fast one-step-ahead cross-validation using a fixed SARIMA model (no repeated auto.arima calls).

Together, these utilities enable fast inspection, diagnostics, and forecasting for all assets without rewriting code for each analysis.


:::

## Univariate Time Series Analysis

### Bitcoin

::: {.panel-tabset}

#### 1. Overview & Stationarity
```{r}
#| warning: false
#| message: false
#| 
plot_original_and_diff("Bitcoin")

run_stationarity("Bitcoin")




plot_acf_pacf("Bitcoin")
```

#### 2. ARIMA
```{r}
#| warning: false
#| message: false

run_arima_models("Bitcoin")



run_arima_diagnostics("Bitcoin")



forecast_arima_asset("Bitcoin")
```

#### 3. SARIMA & Seasonality
```{r}
#| warning: false
#| message: false

run_sarima_models("Bitcoin")



forecast_sarima_asset("Bitcoin")






seasonality_analysis("Bitcoin")
```

#### 4. Benchmarks & Cross-Validation
```{r}
#| results: 'markup'
#| echo: false
#| message: false
#| warning: false


compare_benchmarks("Bitcoin")

```

```{r}

cv_btc <- cross_validate_sarima("Bitcoin")

cat(
  "SARIMA Cross-Validation (Bitcoin)\n",
  "RMSE:", format(round(cv_btc, 6), nsmall = 6), "\n"
)

```

:::



### S&P 500

::: {.panel-tabset}

#### 1. Overview & Stationarity
```{r}
#| warning: false
#| message: false

plot_original_and_diff("SP500")



run_stationarity("SP500")





plot_acf_pacf("SP500")
```

#### 2. ARIMA
```{r}
#| warning: false
#| message: false

run_arima_models("SP500")





run_arima_diagnostics("SP500")


forecast_arima_asset("SP500")
```



#### 3. SARIMA & Seasonality
```{r}
#| warning: false
#| message: false

run_sarima_models("SP500")



forecast_sarima_asset("SP500")



seasonality_analysis("SP500")
```


#### Benchmarks & Cross-Validation
```{r}
#| results: 'markup'
#| echo: false
#| message: false
#| warning: false

compare_benchmarks("SP500")
```

```{r}
cv_spx <- cross_validate_sarima("SP500")

cat(
  " SARIMA Cross-Validation (S&P 500)\n",
  "RMSE:", format(round(cv_spx, 6), nsmall = 6), "\n"
)

```
:::

### NASDAQ
::: {.panel-tabset}

#### 1. Overview & Stationarity
```{r}
#| warning: false
#| message: false

plot_original_and_diff("NASDAQ")



run_stationarity("NASDAQ")



plot_acf_pacf("NASDAQ")
```

#### 2. ARIMA
```{r}
#| warning: false
#| message: false


run_arima_models("NASDAQ")



run_arima_diagnostics("NASDAQ")

#| warning: false
#| message: false

forecast_arima_asset("NASDAQ")
```


#### 3. SARIMA & Seasonality
```{r}
#| warning: false
#| message: false

run_sarima_models("NASDAQ")


forecast_sarima_asset("NASDAQ")



seasonality_analysis("NASDAQ")
```

#### Benchmarks & Cross-Validation
```{r}
#| results: 'markup'
#| echo: false
#| message: false
#| warning: false
#| 
compare_benchmarks("NASDAQ")
```
```{r}
cv_nasdaq <- cross_validate_sarima("NASDAQ")

cat(
  "SARIMA Cross-Validation (NASDAQ)\n",
  "RMSE:", format(round(cv_nasdaq, 6), nsmall = 6), "\n"
)

```
:::


### VIX
::: {.panel-tabset}

#### 1. Time Series + First Difference
```{r}
#| warning: false
#| message: false

plot_original_and_diff("VIX")



run_stationarity("VIX")


plot_acf_pacf("VIX")
```

#### 2. ARIMA
```{r}
#| warning: false
#| message: false

run_arima_models("VIX")

run_arima_diagnostics("VIX")


forecast_arima_asset("VIX")
```



#### 3. SARIMA & Seasonality
```{r}
#| warning: false
#| message: false
run_sarima_models("VIX")

#| warning: false
#| message: false
forecast_sarima_asset("VIX")

#| warning: false
#| message: false
seasonality_analysis("VIX")
```

#### Benchmarks & Cross-Validation
```{r}
#| results: 'markup'
#| echo: false
#| message: false
#| warning: false

compare_benchmarks("VIX")
```
```{r}
cv_vix <- cross_validate_sarima("VIX")

cat(
  " SARIMA Cross-Validation (VIX)\n",
  "RMSE:", format(round(cv_vix, 6), nsmall = 6), "\n"
)


```
:::


### USD
::: {.panel-tabset}

#### 1. Time Series + First Difference
```{r}
#| warning: false
#| message: false
plot_original_and_diff("USD")

#| warning: false
#| message: false
run_stationarity("USD")

#| warning: false
#| message: false
plot_acf_pacf("USD")
```

#### 2. ARIMA
```{r}
#| warning: false
#| message: false
run_arima_models("USD")

#| warning: false
#| message: false
run_arima_diagnostics("USD")

#| warning: false
#| message: false
forecast_arima_asset("USD")
```


#### 3. SARIMA & Seasonality
```{r}
#| warning: false
#| message: false
run_sarima_models("USD")


forecast_sarima_asset("USD")


seasonality_analysis("USD")
```

#### 4. Benchmarks & Cross-Validation
```{r}
#| results: 'markup'
#| echo: false
#| message: false
#| warning: false

compare_benchmarks("USD")
```
```{r}
cv_usd <- cross_validate_sarima("USD")

cat(
  "SARIMA Cross-Validation (USD)\n",
  "RMSE:", format(round(cv_usd, 6), nsmall = 6), "\n"
)


```

:::

All return series show minimal autocorrelation, consistent with financial market efficiency. Bitcoin fits a simple ARMA(1,0), SPX fits ARMA(2,1), VIX requires a more complex ARMA(3,4), and USD behaves like pure noise. In short, daily returns are almost unpredictable, and models capture only very small short-term dynamics.

